---
title: "Descriptive Statistics"
author: "Huilin Lu"
date: "12/9/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

### Using Base R to Generate Statistical Indicators

```{r eval = TRUE}
demo <- read.csv("demographics.csv")

##########
### how to compute the main statistical indicators
### for a numeric variable in base R
##########
####### we will compute these indicators for the variable income
### mean
mean(demo$income)
### or
m <- mean(demo$income)
print(m)
### standard deviation and variance
sd(demo$income)
var(demo$income)
### minimum, maximum and range
min(demo$income)
max(demo$income)
range(demo$income)
max(demo$income) - min(demo$income)
### median
median(demo$income)
### quartiles, 0%, 25%, 50%, 75%, 100%
quantile(demo$income)
```

### Descriptive Statistics with the Psych Packgae

```{r eval = TRUE}
##########
### how to compute the main statistical indicators for a numeric variable
### with the psych package
##########
###### we will compute this indicators for the following variables
###### age, income and car price
### create a matrix with the variables of interest
demo2 <- cbind(demo$age, demo$income, demo$carpr)
### give suggestive names to the matrix columns
## Matrix can not use '$', need to convert to dataframe
colnames(demo2) <- c("age", "income", "price")
head(demo2)
### load the psych package
library(psych)

### use the describe function to generate the statistics table
describe(demo2)
### the trimmed mean is computed with a default trim of 0.1
### mad - median absolute deviation (the median of the absolute deviations from the data median)
######## more options for the describe function
describe(demo2, na.rm = TRUE, trim = 0.1, check = TRUE)
### na.rm - if TRUE it omits the missing values (if FALSE it includes the cases when calculatiing)
### trim - sets the trimming fraction, exclude data points % from top and bottom
### check - if TRUE it checks for non-numeric data

```


### Descriptive Statistis with the pastecs Package

```{r eval = TRUE}
### load the pastecs package
library(pastecs)

### before computing the indicators we set some options (in base R)
options(scipen=100) ## force R to use the standard notation, NOT the exponential notation
options(digits=2) ## make R show only the first two decimals
### run the stat.desc funtion from pastecs
### if we want ALL the statistics we run
stat.desc(demo2)
## We have essential statistics: CI.mean.0.95 (95% upper bound of mean), coefficient of variation (standard deviation / mean)
### if we want to omit the basic statistics we run
stat.desc(demo2, basic = FALSE)

### if we want the basic statistics only we can execute
stat.desc(demo2, desc = FALSE)   ### The opposite of basic = FALSE
```



### Determining the Skewness and Kurtosis

```{r eval = TRUE}

##########
### how to compute skewness and kurtosis
### with the e1071 package
##########
### we will use the variable income for our examples
### load the package
library(e1071)
### compute the skewness: positive skewness: Right Skewed
skewness(demo$income)

### compute the kurtosis: positive kurtosis: peak 
kurtosis(demo$income)

```



### Computing Quantiles

```{r eval = TRUE}


##########
### how to compute quantiles
##########
### we will use the variable income for our example
### compute the following percentiles
### 17%, 55% and 97%
### use the quantile function in the stats package
### (this package loads automatically when you start R)
quantile(demo$income, probs = c(0.17, 0.55, 0.97))
### to get the quartiles
quantile(demo$income, probs = c(0.25, 0.50, 0.75))



```


### Determining the Mode

```{r eval = TRUE}

##########
### how to determine the mode of a numeric variable
### with the package modeest
##########
### we will find out the mode for the variable income
### load the package
library(modeest)
mlv(demo$income, method="mfv") ### "mfv" stands for "most frequent value"
### for the discrete variables, the best way to compute the mode is to tabulate the frequencies
### as we will learn in a future lecture of this course


```


### Getting the Statistical Indicators by Group with doBy

```{r eval = TRUE}

##########
### how to compute the main statistical indicators by groups or subsets
### with the package doBy
##########
### we will get the main statistical indicators for the variable income
### by gender (separately for male and female subjects)

### load the package
library(doBy)
### define the function that generates the statistics
### N.B. descStat is the command (in the doBy package) that computes the statistical indicators
func <- function(x) {descStat(x, na.rm=TRUE)}
### use the command summaryBy
summaryBy(income~gender, data=demo, FUN=func)
### get the main statistical indicators for the variables income and age
### by gender
summaryBy(income+age~gender, data=demo, FUN=func)

```


### Getting the Statistical Indicators by Group with describeBy  (psych library)


```{r eval = TRUE}
##########
### how to compute the main statistical indicators by groups or subsets
### with the package psych (command describeBy)
##########
### we will get the main statistical indicators for the variable income
### by education level
### load the package
library(psych)
### use the describeBy command
describeBy(demo$income, demo$educ)

### The disadvantage of descibeBy() function is that we can not have multiple variables calculating simultaneously,
### we need to use summaryBy() function in doBy instead
```




### Getting the Statistical Indicators by Group with Stats.


```{r eval = TRUE}

##########
### how to compute the main statistical indicators by groups or subsets
### with the package stats (command aggregate)
##########
### we will get some statistical indicators for the variable age
### by marital status (married/unmarried)
### compute the mean
aggregate(demo$age, by=list(demo$marital), FUN=mean)
### compute the standard deviation
aggregate(demo$age, by=list(demo$marital), FUN=sd)
### compute the median
aggregate(demo$age, by=list(demo$marital), FUN=median)
### compute the variance
aggregate(demo$age, by=list(demo$marital), FUN=var)
### etc.
### very useful when we want to combine the factor categories
aggregate(demo$agem by = list(demo$marital, demo$gender), FUN = mean)

```













